---
title: "Regional Scenarios"
author: ""
date: "2022/07/28"
output: html_document
---

```{r, Setup, message=FALSE, warning=FALSE,fig.width=12, fig.height=12}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(reshape2)
library(RColorBrewer)
theme_set(theme_light(base_size = 18))

# set up, colour palettes etc.
age_lvls <- c("0-4","5-11","12-17","18-39","40-59","60-79","80+")
ha_lvls <- c("Fraser", "Interior", "Vancouver Island", "Vancouver Coastal", "Northern")
age_pal <- "YlOrRd"
ha_pal <-c("Fraser"="#3B9AB2","Interior"="#78B7C5", "Vancouver Island"="#EBCC2A", "Vancouver Coastal"="#E1AF00", "Northern"="#F21A00")

data_dir <- "../data/"
output_dir <- "./"  # generate: infections by age and HA, immunity profiles by age and HA and wave
```  
## Vaccine Data

Load in and look at vaccination data by age and HA (separately):
  
```{r,message=FALSE, warning=FALSE,fig.width=12, fig.height=12}
# Vaccinations by HA
vaxdat <- readr::read_csv(paste0(data_dir,"HA_cumulative_2022-07-07.csv")) %>%
  mutate(date=mdy(date)) %>%
  melt(id=c("date","age_group","region", "population")) %>%
  dplyr::select(-c("age_group")) %>%
  filter(!(grepl("total", variable)))  %>%
  group_by(date=floor_date(date, "month"), region,variable) %>%
  summarize(value=max(value)) 

vaxdat$region <- factor(vaxdat$region, levels=ha_lvls)

ggplot(vaxdat,aes(x=date, y=value, col=region))+
  geom_line(size=2)+
  facet_wrap(~variable)+
  scale_color_manual(values=ha_pal)+
  labs(x="", y="Vaccine Coverage", col="Health Authority")

HApopdat <- readr::read_csv(paste0(data_dir,"HA_cumulative_2022-07-07.csv"))  %>%
  group_by(region) %>%
  summarize(population)
HApopdat <- HApopdat[!duplicated(HApopdat),] # Note: not by age

# Vaccinations by age
vaxdat2 <- readr::read_csv(paste0(data_dir, "BC_cumulative_2022-07-07.csv")) %>%
  mutate(age_group=case_when(age_group=="17-Dec"~"12-17", #miscoded 
      age_group=="11-May"~"5-11",
      TRUE ~ age_group)) %>%
  mutate(date=mdy(date)) %>%
  filter(age_group != "5+", age_group != "12+",age_group != "18+",
    age_group != "50+",age_group != "All")  %>%
  group_by(date=floor_date(date, "month"), age_group, dose_number) %>%
  summarize(coverage=max(coverage))


ggplot(vaxdat2, aes(x=date, y=coverage, col=age_group))+
  geom_line(size=2)+
  facet_wrap(~dose_number, labeller=label_both)+
  scale_color_brewer(palette=age_pal)+
  labs(x="", y="Vaccine Coverage", col="Age")

```
  
## Case Data
Load in and look at reported cases by age and HA (combined):
  
```{r,message=FALSE, warning=FALSE,fig.width=12, fig.height=12}
  
# Cases by age and HA 
casedat <- readr::read_csv(paste0(data_dir, "BCCDC_COVID19_Dashboard_Case_Details.csv")) %>%
  dplyr::select(date=Reported_Date,
    age_group=Age_Group,
    region=HA) %>%
  mutate(date=ymd(date)) %>%
  filter(region %in% ha_lvls) %>%
  group_by(age_group, region, date) %>%
  summarize(cases=n()) %>%
  filter(age_group != "Unknown")# %>%
 # group_by(date=floor_date(date, "month"), region, age_group) %>%
 # summarize(cases=sum(cases))

casedat$region <- factor(casedat$region, levels=ha_lvls)

ggplot(casedat,aes(x=date, y=cases, fill=age_group))+
  geom_area()+
  facet_wrap(~region, scales="free_y")+
  scale_fill_brewer(palette=age_pal)+
  labs(x="", y="Monthly Reported Cases", fill="Age")


```

## The fudging begins.

We'll need to combine some age groups to get consistent data. Let's go with: 0-29 :( , 30-69, 70+.

```{r,message=FALSE, fig.width=12, fig.height=12}


vaxdat2 <- vaxdat2 %>%
   mutate(age_group=case_when(
    age_group %in% c("5-11","12-17", "18-29") ~ "0-29",
    age_group == "70+" ~ "70+",
    TRUE ~ "30-69")) %>%
  group_by(age_group, date, dose_number) %>%
  summarize(coverage=mean(coverage)) # average for now..should use demographics


casedat <- casedat %>%
  mutate(age_group=case_when(
    age_group %in% c("<10","10-19", "20-29") ~ "0-29",
    age_group %in% c("70-79","80-89","90+") ~ "70+",
    TRUE ~ "30-69")) %>%
  group_by(age_group, date, region) %>%
  summarize(cases=sum(cases))


```


Get some age demographics. Assume same demographics for each HA...
 
source: https://bcstats.shinyapps.io/popApp/
```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
N <- 5000880
BC_demo <- data.frame(
  age_group=c("0-29", "30-69", "70+"),
  p=c((716900+253690+294650+337090)/N,
      (3267620-(253690+294650+337090)+319405)/N,
      (1016365-319405)/N)) # note: get rid of this now?

HA_demo <- readr::read_csv(paste0(data_dir, "Population_Projections_HA.csv")) %>%
  dplyr::select(region=`Health Authority`,
    '0-29' = `0 to 29`, '30-69'=`30 to 69`,
    '70+' = `70 to 90+`) %>%
  melt(id="region") %>%
  rename(age_group=variable, population=value)

```


CC's clumsy ascertainment fractions, using the demographics from stats canada but with more current total population of BC, CITF serology data, BCCDC serology data and cases in daily rather than monthly groups.  

```{r, message=FALSE} 
casesbc <- readr::read_csv(paste0(data_dir, "BCCDC_COVID19_Dashboard_Case_Details.csv")) %>%
  dplyr::select(date=Reported_Date,
    age_group=Age_Group,
    region=HA) %>%
  mutate(date=ymd(date)) %>%
  filter(region %in% ha_lvls) %>%
  group_by(age_group, region, date) %>%
  summarize(cases=n()) %>%
  filter(age_group != "Unknown") %>%
  group_by(date, region, age_group) %>%
  summarize(cases=sum(cases)) %>%
  mutate(age_group=case_when(
    age_group %in% c("<10","10-19", "20-29") ~ "0-29",
    age_group %in% c("70-79","80-89","90+") ~ "70+",
    TRUE ~ "30-69")) %>%
  group_by(age_group, date, region) %>%
  summarize(cases=sum(cases))  %>% 
  ungroup() %>% group_by(date, age_group) %>% 
  summarise(cases = sum(cases))

# pre-omicron
preom = casesbc %>% filter(date< ymd("2021-10-01")) %>% group_by(age_group) %>% 
  summarize(total = sum(cases)) %>% mutate(frac.reported = total/(BC_demo$p*5.07e6), frac.serology = c(0.1,0.09, 0.06)) %>% mutate(UAR = frac.serology/frac.reported,period="pre-omicron")

# just after ba1
ba1asc =  casesbc %>% filter(date< ymd("2022-03-30") & date >= ymd("2021-10-01")) %>% group_by(age_group) %>% 
  summarize(total = sum(cases)) %>% mutate(frac.reported = total/(BC_demo$p*5.07e6), frac.serology = c(0.58,0.4, 0.14) - preom$frac.serology) %>% mutate(UAR = frac.serology/frac.reported,period="ba1")
```

Now we try to use CITF data for serology post-BA2. This is quite fudged because children and presumably the elderly don't donate blood, and CITF for March 2022 disagrees with BCCDC, with the BCCDC estimate at overall 43% of BC with infection, vs CITF at 32%. 

We know that testing continued to change, because of the hospitalization and wastewater signals and the CITF serology numbers, so we can't just use reported cases. We could add a sanity check with reported cases in those over 70. 

CITF gets BA2 taking BC from 32% to 45% March to May, reflecting an increase by 12% of the population. If this is underestimate by a factor of (43/32) (difference in March) then BC's adjusted May estimate would be 43+12(43/32)=59% (mid-May). We will have to invent how this was distributed across ages. In the CITF data the biggest increase was in those over 60, and somewhat smaller increases happened in the other groups. 


```{r, message=FALSE} 
ba2asc =  casesbc %>% filter(date< ymd("2022-05-15") & date >= ymd("2022-03-30")) %>% group_by(age_group) %>% 
  summarize(total = sum(cases)) %>% mutate(frac.reported = total/(BC_demo$p*5.07e6), total.under.ba1 = total*ba1asc$UAR)
sum(ba2asc$total.under.ba1)/N
```

If we assumed that the ascertainment stayed the same, only 2% of BC would have had BA2, but serology from CITF, wastewater, hospitalizations and cases in those over 70 suggest that the BA2 peak was about half as high as the BA1 peak. I am therefore going to assume that the UAR in the over 70s remained nearly what it was under BA1 and choose the other two groups to match the serology. I set it so that 14% of BC got BA2. This is between the CITF 12% and what I would guess if I took that 12% and adjusted for the difference between BCCDC and CITF (a factor of 43/32, making it 16%). 

```{r, message=FALSE} 
ba2asc =  casesbc %>% filter(date< ymd("2022-05-15") & date >= ymd("2022-03-31")) %>% group_by(age_group) %>% 
  summarize(total = sum(cases)) %>% mutate(frac.reported = total/(BC_demo$p*5.07e6), UAR=c(8,9,1.5)*ba1asc$UAR, period = "ba2") %>% mutate(frac.serology=UAR*total)
ba2asc
```

Now we include these estimates of the UAR into the corrected cases. Before Dec 25, use pre-Omicron numbers. Use BA1 numbers until April 1, 2022, and BA2 numbers thereafter. 

```{r, message=FALSE, fig.width=12, fig.height=12}
casedat <- casedat %>%
  mutate(corrCases=case_when(
    date < ymd("2021-12-25") & age_group == "0-29" ~ cases*preom$UAR[1],
    date < ymd("2021-12-25") & age_group == "30-69" ~ cases*preom$UAR[2],
    date < ymd("2021-12-25") & age_group == "70+" ~ cases*preom$UAR[3],
    date < ymd("2022-04-01") & date >= ymd("2021-12-25") & 
      age_group == "0-29" ~ cases*ba1asc$UAR[1],
    date < ymd("2022-04-01") & date >= ymd("2021-12-25") & 
      age_group == "30-69" ~ cases*ba1asc$UAR[2],
    date < ymd("2022-04-01")  & date >= ymd("2021-12-25")& 
      age_group == "70+" ~ cases*ba1asc$UAR[3],
    date >= ymd("2022-04-01") & age_group == "0-29" ~ cases*ba2asc$UAR[1],
    date >= ymd("2022-04-01") & age_group == "30-69" ~ cases*ba2asc$UAR[2],
    date >= ymd("2022-04-01") & age_group == "70+" ~ cases*ba2asc$UAR[3]))
```
Look at the corrected cases by age and output a csv:

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
## add demos to case dat
casedat <- casedat %>%
  inner_join(HA_demo) %>%
  mutate(perc_cases= corrCases/population)

# reformat
casedat <- casedat %>%
  dplyr::select(age_group,date,region,value=perc_cases) %>%
  mutate(variable="infections") 

ggplot(casedat,
  aes(x=date, y=value, col=age_group,fill=age_group))+
  geom_col(size=1.5)+
  facet_wrap(~region)+
  scale_colour_brewer(palette=age_pal)+
  scale_fill_brewer(palette=age_pal)+
  labs(x="", y="Estimated Infections (% pop)", fill="Age", col="Age")

# save
readr::write_csv(casedat, paste0(output_dir, "BC_infections_by_age_and_ha.csv"))


# now group cases by month for immunity profiles
casedat <- casedat %>%
  group_by(date=floor_date(date, "month"), variable, region, age_group) %>%
  summarize(value=sum(value))
```


Same kind of output, but for hospitalizations, for help validating the regional model: 

```{r}
agehosp <- readr::read_csv(paste0(data_dir, "all_hosps_byage1.csv")) %>%
  mutate(week_of = case_when(week >= 44 ~ ymd("2021-10-31")+7*(week-44),
                             week < 44 ~ ymd("2022-01-02") + 7*(week-1))) %>%
  mutate(age_group=case_when(
    age_group %in% c("<10","10-19", "20-29") ~ "0-29",
    age_group %in% c("70-79","80-89","90+") ~ "70+",
    TRUE ~ "30-69")) %>%
  group_by(age_group,week_of) %>%
  summarize(admissions=sum(new))

# save
readr::write_csv(agehosp, paste0(output_dir, "bc-admissions-by-age.csv"))
```

And again, now with the data that has both HA and age in it (the previous block only has BC-level data in it : 

```{r}
hdat <- readr::read_csv(paste0(data_dir, "5 Aug 2022_ha.csv")) %>%
  dplyr::select(date=week,
                age_group=age_group,
                region=ha_name,
                hosp=nhosps_week_grouped) %>%
  mutate(date=ymd(date)) %>% 
  mutate(age_group=case_when(
    age_group %in% c("5-17","18-29") ~ "0-29",
    age_group %in% c(">=70") ~ "70+",
    TRUE ~ "30-69")) %>% group_by(date, age_group, region) %>% summarize(hosp = sum(hosp))

glimpse(hdat)
ggplot(hdat,aes(x=date, y=hosp, fill=age_group))+
  geom_area()+
  facet_wrap(~region, scales="free_y")+
  scale_fill_brewer(palette=age_pal)+
  labs(x="", y="Weekly admissions", fill="Age")

write_csv(hdat, file = paste0(output_dir,"admission-age-ha-3bands.csv"))
```


OK, now estimate vaccination coverage by both age and HA. Assuming the provincial
age trends hold across the province..
  
```{r, message=FALSE, warning=FALSE,fig.width=12, fig,height=12}

vaxdat <- vaxdat %>%
  mutate(dose_number=tidyr::extract_numeric(variable)) %>%
  inner_join(vaxdat2) %>%
  mutate(value=value/100, coverage=coverage/100) %>%
  mutate(combined_coverage=coverage*value)

```
# Building Immunity Profiles

```{r, message=FALSE, fig.width=12, fig.height=12}
# first get new vaccinations by age and HA
vaxdat_diff <- vaxdat %>%
  dplyr::select(date, region,age_group, variable, value=combined_coverage) %>%
  group_by(variable,age_group,region) %>%
  nest() %>%
  mutate(value=map(data, function(x) {
    data.frame(value=diff(x$value,1),
      date=x$date[-1])})) %>%
  unnest(value) %>%
  dplyr::select(-c(data)) %>%
  group_by(age_group,region) %>% 
  tidyr::complete(variable, 
      date=seq(min(vaxdat$date), max(vaxdat$date), by="month"),
      fill=list(value=0)) %>% #add missing dose_4
  filter(!(grepl("total", variable))) 


vaxdat_diff$variable <- sub("_cumulative_coverage", "", vaxdat_diff$variable)

ggplot(vaxdat_diff, aes(x=date, y=value, fill=variable))+
  geom_col()+facet_grid(age_group~region)+
  labs(x="", y="New vaccinations as % pop", fill="")+
  theme(axis.text.x = element_text(angle =65,vjust=0.1))

```
Try to deal with overcounting. For each (n > 1)th dose, remove a (n-1)th dose that occurred earlier in the data..

```{r, message=FALSE, fig.width=12, fig.height=12}
remove_early_doses <- function(dat){ 
    ####################
    newdat <- data.frame()
    for (n in c(1,2,3)){
      dat1 <- filter(dat, variable==paste0("dose_",n)) %>% arrange(date)
      dat2 <- filter(dat, variable==paste0("dose_",n+1)) %>% arrange(date)
      v1 <- dat1$value
      s <- min(which(v1 > 0))
      for (v2 in dat2$value){
        if (v2 > 0) {
          # start with oldest dose 1s to remove
          v1[s] <- v1[s]-v2
          # now bounce:
          for (i in s:length(v1)){
            if (v1[i] < 0) {
              v1[i+1] <- v1[i+1] + v1[i]
              v1[i] <- 0
            }
          }
        }
      }
      dat1$value <- v1
      # update
      newdat <- rbind(newdat,dat1)
    } 
    newdat <- rbind(newdat, filter(dat, variable=="dose_4"))
  return(newdat)
  }


```
  
## Regional immunity profiling model

```{r}
# Model defn
# start date means start of wave (so look at all
# date up to that point)
immunity_model <- function(vax_dat,case_dat,start_date){
  # Model variables
  w = 0.05 # waning fraction per month
  Vi_vax_0 = 0.8 # initial protection from vax
  Vi_pi_0 = 0.8 # initial protection from infection

  # first remove over-counted vax doses
  # note: important to do this after setting start_date
  vax_dat <- vax_dat %>%
    filter(date <= start_date)%>%
    group_by(age_group, region) %>%
    nest() %>%
    mutate(data=map(data, remove_early_doses))%>%
    unnest(cols=c(data))

  # combine case dat
  immunity_dat <- rbind(
    filter(case_dat, date <= start_date),
    vax_dat)

  # calculate profiles
  res <- immunity_dat %>%
    mutate(time = interval(date, start_date) %/% months(1)) %>% 
    pivot_wider(names_from=variable, values_from=value) %>%
    mutate(p_pi=infections, p_vax=dose_1+dose_2+dose_3+dose_4) %>% 
    mutate(Vi_vax=p_vax*Vi_vax_0*(1-w)^time,
          Vi_pi=p_pi*Vi_pi_0*(1-w)^time) %>%
    dplyr::select(c(region, age_group, Vi_vax, Vi_pi)) %>%
    group_by(region,age_group) %>%
    summarize(Vi_pess= 1-(1-sum(Vi_vax, na.rm=TRUE))*(1-sum(Vi_pi, na.rm=TRUE)),
      Vi_opt = min(1, sum(Vi_vax, na.rm=TRUE)+sum(Vi_pi, na.rm=TRUE)),
      Vi = (Vi_pess+Vi_opt)/2)

  return(res)
}

```
## pre-BA.1 immunity

```{r, message=FALSE, warning=FALSE}
BA1_start_date <- ymd("2021-12-01")
summ_immunity_BA1 <- immunity_model(vaxdat_diff,casedat, BA1_start_date)

# save
readr::write_csv(summ_immunity_BA1, paste0(output_dir, "BC_protection_BA1.csv"))

# look at results
summ_immunity_BA1 %>%
  knitr::kable()

```
## pre-BA.2 immunity

```{r, message=FALSE}
BA2_start_date <- ymd("2022-03-01")
summ_immunity_BA2 <- immunity_model(vaxdat_diff,casedat, BA2_start_date)

# save
readr::write_csv(summ_immunity_BA2, paste0(output_dir, "BC_protection_BA2.csv"))

summ_immunity_BA2 %>%
  knitr::kable()

```
## pre-BA.5 immunity

```{r, message=FALSE}
BA5_start_date <- ymd("2022-06-01")
summ_immunity_BA5 <- immunity_model(vaxdat_diff,casedat, BA5_start_date)

# save
readr::write_csv(summ_immunity_BA5, paste0(output_dir, "BC_protection_BA5.csv"))

summ_immunity_BA5 %>%
  knitr::kable()


```

# Contact Matrix
We use the Canadian contact matrix estimated in [Prem et al.](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005697#sec014) from POLYMOD data.

This matrix has entries $\mu_{a,i}$, the total number of contacts an age-$a$ individual makes with an age-$i$ individual per day.
```{r, message=FALSE}
# Load in contact matrix for canada
contact_df <- readr::read_csv(paste0(data_dir, "MUestimates_all_locations_CAN.csv"))
contact_mat <- as.matrix(contact_df)
rownames(contact_mat) <- colnames(contact_mat)

heatmap(contact_mat, Rowv=NA, Colv=NA)

```
  
In an age-structured SIR model, we would have for each age group $a$,
$$\frac{dS_a}{dt} = -\beta S_a \sum_{i} \mu_{a,i} \frac{I_i}{N_i}.$$
i.e, the contact rate from age group $i$ to age group $a$ is 
$$\frac{\mu_{a,i}}{N_i}.$$ 
So for new age groups $a_1=i+j$, $a_2=k+l$, we take
$$\mu_{a_1,a_2} = \frac{p_i \left(\mu_{i,k}+\mu_{i,l}\right)+p_j \left(\mu_{j,k}+\mu_{j,l}\right)}{p_i+p_j},$$
where $p_i$ is $N_i/N$. So note in particular that we need demographic data for the original age bands.

```{r}
# load pop data in 5yr bands
HA_demo_by_fives <- readr::read_csv(paste0(data_dir, "Population_Projections_5yr_bands.csv"))
# put into same bands as contact matrix
HA_demo_by_fives <- HA_demo_by_fives %>%
  filter(Gender=="T") %>%
  mutate(`0 to 4` = LT1+`1 to 4`, `75+` = `75 to 79` + `80 to 84` + `85 to 89` + `90+`) %>%
  dplyr::select(-c(Region,Year, Gender, LT1, `1 to 4`,`75 to 79`,`80 to 84`,`85 to 89`,`90+`)) %>%
  group_by(`Health Authority`) %>%
  mutate(across(where(is.numeric), ~ .x/Total))%>%
  dplyr::select(-c(Total)) %>% rename(region=`Health Authority`)

HA_demo_by_fives <- HA_demo_by_fives[, c('region', rownames(contact_mat))] # reorder to match

breaks <- c(0, 6, 14, 16) # combine grps 1-6, 7-14, and 14-16
names <- c('0-29', '30-69', '70+')
for (ha in ha_lvls) {
    nrow = length(breaks)-1
    new_contact_mat <- matrix(0.0, ncol=nrow, nrow=nrow)
    pop_dat <- filter(HA_demo_by_fives, region == ha) %>% ungroup() %>% dplyr::select(-c(region))
    for (r in 1:nrow) {
        start_r <- breaks[r]+1
        end_r <- breaks[r+1]
        for (c in 1:nrow) {
            ss <- 0.0
            start_c <- breaks[c]+1
            end_c <- breaks[c+1]
            for (i in start_r:end_r){# look over original matrix
              ss <- ss + pop_dat[[i]]*(sum(contact_mat[i,start_c:end_c]))
            }
            new_contact_mat[r,c] <- ss/sum(pop_dat[start_r:end_r])         
        }
      # save
      fname <- paste0("contactmat_newages_", ha, ".csv")
      rownames(new_contact_mat) <- colnames(new_contact_mat) <- names
      write.csv(new_contact_mat, paste0(output_dir, fname))
   }
}

```



